Stellar Bars in Isolated Gas-Rich Spiral Galaxies Do Not Slow Down

### 2022-04-23 paper/ms.tex
Change title
Tom Dame edits
And other small things I noticed

\title{Stellar Bars in Isolated Gas-Rich Galaxies Do Not Slow Down}

@@

  Elongated bar-like features are ubiquitous, occuring at the centres of
  approximately two-thirds of spiral disk galaxies \cite{2000AJ....119..536E,
  2007ApJ...657..790M}. Due to gravitational interactions between the bar and
  the other components of galaxies, it is expected that angular momentum and
  matter will redistribute between galactic components over long (Gyr)
  timescales in galaxies hosting a bar \cite{1972MNRAS.157....1L,
  1984MNRAS.209..729T, 1985MNRAS.213..451W}. Previous work has overwhelmingly
  provided the expectation that, due to these interactions, the bar pattern
  will slow its rotation over time \cite{1992ApJ...400...80H,
  2000ApJ...543..704D, 2002MNRAS.330...35A, 2002ApJ...569L..83A,
  2003MNRAS.341.1179A, 2003MNRAS.346..251O, 2005MNRAS.363..991H,
  2006ApJ...637..214M, 2007MNRAS.375..460W, 2009ApJ...697..293D}. We have
  performed a simulation of an isolated galactic disk hosting a strong bar
  which includes a state-of-the-art model of the interstellar medium. Here we
  show that in this simulation the bar pattern does not slow down over time,
  and instead remains at a stable, constant rate of rotation. This behavior
  has been observed in previous simulations but its explanation has remained
  elusive.\cite{1993AA...268...65F, 2010ApJ...719.1470V}. We propose an
  equilibrium mechanism which is consistent with our simulations. This result
  challenges our expectations for how barred galaxies redistribute angular
  momentum and matter over Gyr timescales, which is especially relevant for
  our understanding of how the Milky Way arrived at its present day state.

@@

It has long been known that non-axisymmetric features in a stellar disk act to
redistribute mass and angular momentum.\cite{1972MNRAS.157....1L} The
interaction between a bar and a spheroid (either a dark matter halo or stellar
bulge) has received considerable interest.\cite{1984MNRAS.209..729T,
1985MNRAS.213..451W} A reasonable stellar disk in isolation seems to always
generate a bar instability,\cite{1971ApJ...168..343H} but the presence of a
static dark matter halo acts to stabilize the disk against the bar
instability.\cite{1973ApJ...186..467O, 1976AJ.....81...30H} Later, a number of
numerical simulations have confirmed that a live dark matter halo generically
exerts a negative torque on the bar, causing its pattern speed to decrease and
its length and strength to increase\cite{1992ApJ...400...80H,
2000ApJ...543..704D, 2002MNRAS.330...35A, 2002ApJ...569L..83A,
2003MNRAS.341.1179A, 2003MNRAS.346..251O, 2005MNRAS.363..991H,
2006ApJ...637..214M, 2007MNRAS.375..460W, 2009ApJ...697..293D}. Intuitively,
the bar excites a wake of resonant material in the dark matter halo in a
manner similar to dynamical friction. This wake lags and thus exerts a
negative torque on the bar.

The pre-eminent expectation that the dark matter halo acts to slow the bar
down conflicts with observational estimates of bar pattern speeds. Bar
rotation rates are typically classified by the dimensionless ratio
\begin{equation}
\Rot = \RCR/R_b\textrm{,}
\end{equation}
where \RCR{} is the radius of corotation and $R_b$ is the length of the
bar.\footnote{The radius of corotation \RCR{} is defined for circular orbits as
the radius at which the orbital frequency is equal to the pattern speed,
$\Omega_p$, of a given non-axisymmetric feature. In a galaxy with a constant
circular velocity $V_c$, it is given by $\RCR = V_c / \Omega_p$.} Galaxies
with $\Rot < 1.4$ are considered ``fast rotators'' while galaxies with $\Rot >
1.4$ are considered ``slow rotators.''\cite{2000ApJ...543..704D} Galaxies with
$\Rot < 1$ are not thought to be stable.\cite{1980AA....81..198C}
Observational estimates of the pattern speeds of bars indicate that nearly all
galaxies have $1 < \Rot < 1.4$.\cite{2011MSAIS..18...23C, 2015AA...576A.102A,
2019MNRAS.482.1733G}

@@

A surface density projection of our simulatons is shown in
Fig.~\ref{fig:overview}. The upper three panels show the disk in the N-body
run while the lower three panels show the disk in the SMUGGLE run. Each column
shows the disk $\sim1\,\textrm{Gyr}$ apart in time. There is a large qualitative
difference in the evolution of the bar pattern between the two runs. We see
that in the N-body case, the bar lengthens in time and grows in strength. In
the SMUGGLE case, the bar pattern retains a similar length and strength
between panels.

@@

We can test this interpration by measuring the angle offset between the halo
wake and the bar. If the wake and the bar are aligned (i.e., there is no angle
offset), then the wake cannot exert a negative torque on the bar. This angle
is plotted in the middle panel of Fig.~\ref{fig:wake}, which shows that the
angle offset is larger in the N-body case than in the SMUGGLE case by about a
factor of two. The left and right panels of Fig.~\ref{fig:wake} show the halo
wake with respect to the location of the bar in the N-body (left) and SMUGGLE
(right) cases.

@@

When the bar slows down, we argue that this induces a larger positive torque
from the gas phase. Only gas within corotation will flow inwards, while gas
outside corotation will flow outwards.\cite{2011MNRAS.415.1027H} Since the
corotation radius is larger for more slowly rotating bars, it follows that
more slowly rotating bars should be more efficient at driving gas inflows and
thus experience a larger positive torque from the gas phase. We performed an
experiment to test this hypothesis by forcing the stellar disk in the SMUGGLE
run to rotate at a constant angular rate and measuring the torque on the bar
by the gas phase at different rotation rates. The result of this experiment is
shown in Fig.~\ref{fig:equil}, which shows that a more slowly rotating bar
experiences a larger positive torque from the gas.

We have therefore shown evidence for an equilibrium mechanism at play which
keeps the pattern speed of the bar constant, resulting from the complex
interplay between the dark matter halo and the gas phase.

@@

Barred galaxies in cosmological simulations of galaxy formation continue to be
in conflict with observations.\cite{2017MNRAS.469.1054A, 2019MNRAS.483.2721P,
2021AA...650L..16F} However, the pattern speeds of bars in both cosmological
simulations and the real universe can be affected by environmental processes
not included in our simulation -- e.g., satellite
infall\cite{2011Natur.477..301P}, non-sphericity\cite{2013MNRAS.429.1949A} or
rotation\cite{2013MNRAS.434.1287S, 2014ApJ...783L..18L, 2018MNRAS.476.1331C,
2019MNRAS.488.5788C} in the dark matter halo, or perhaps even the gaseous
circumgalactic medium. Naturally, extending our present work to account for
such effects is a crucial next step in understanding the formation and
evolution of galactic bars.

@@

\begin{figure}[h]%
\centering
\includegraphics[width=0.45\textwidth]{fig/fig2.pdf}
\caption{Various bar properties from the N-body and SMUGGLE runs. The
\textit{upper panel} show the evolution of the pattern speed. As expected, the
bar in the N-body run slows down due to interactions between the bar and the
dark matter halo. However, the bar in the SMUGGLE run does not slow down and
instead remains at a constant pattern speed. The \textit{middle panel} shows
the evolution of the bar length. In the N-body case, the bar lengthens. This
occurs because as the pattern speed drops, bar-like orbits at larger radii are
possible. Stars are captured on these orbits, lengthening the bar. This
process does not occur in the SMUGGLE cases since the bar pattern speed is not
decreasing, and therefore the bar length remains constant. The \textit{lower
panel} shows the torque on the bar by different components. The solid lines
show the torque exerted by the halo in both the N-body and SMUGGLE cases. The
dashed line shows the torque exerted by the gas phase in the SMUGGLE run
(there is no gas in the N-body run). The first $1.5\,\textrm{Gyr}$ of
evolution of the N-body model is not shown. Details on how these properties
are calculated is given in the Methods section.}\label{fig:prop}
\end{figure}

@@

Star formation occurs in cells above a density threshold
($n_{\textrm{th}}=100\,\textrm{cm}^{-3}$) according to SH03 with a
star-formation efficiency of $\epsilon = 0.01$. Star formation converts gas
cells into star particles which represent single stellar populations. For each
star particle, the deposition of energy, momentum, and mass from stellar winds
and supernovae is modeled. Photoionization and radiation pressure are modeled
using an approximate treatment. A more detailed description of this model can
be found in the flagship SMUGGLE paper.\cite{2019MNRAS.489.4233M}

We used the fiducial model parameters, except that we increased the number of
effective neighbors $N_{\textrm{ngb}}$ for the deposition of feedback from
$64$ to $512$. We found that a lower value of $N_{\textrm{ngb}}$ resulted in
inefficient photoionization feedback since the photoionizing budget had not
been exhausted after deposition into $64$ neighboring cells. We also used an
updated version of SMUGGLE using a new mechanical feedback routine similar to
the one described in ref.\cite{2018MNRAS.480..800H} This updated routine is a
tensor renormalization which ensures linear and angular momentum conservation
to machine precision.

@@

After generating the gaseous disk in this way, we stitched the gas disk
together with the GALAKOS N-body disk (and bulge and dark matter halo) after
the GALAKOS disk has been allowed to evolve for $1.5\,\textrm{Gyr}$. The
purpose of allowing the GALAKOS disk to evolve first for a short period of
time is to allow for the bar to form unimpacted by the presence of the gas
(which would normally disrupt the formation of the bar). We made one
additional modification when stitching the gas disk together with the N-body
disk - we created a vacuum within the central $4\,\textrm{kpc}$. This vacuum
guards against an initial dramatic infall of gas within the bar region, which
we found to destroy the bar.

>>>>

\title{Stellar Bars in Isolated Gas-Rich Spiral Galaxies Do Not Slow Down}

@@

  Elongated bar-like features are ubiquitous, occurring at the centers of
  approximately two-thirds of spiral disk galaxies \cite{2000AJ....119..536E,
  2007ApJ...657..790M}. Due to gravitational interactions between the bar and
  the other components of galaxies, it is expected that angular momentum and
  matter will redistribute between galactic components over long (Gyr)
  timescales in galaxies hosting a bar \cite{1972MNRAS.157....1L,
  1984MNRAS.209..729T, 1985MNRAS.213..451W}. Previous work has overwhelmingly
  provided the expectation that, due to the interaction between the bar and the halo, the bar pattern
  will slow its rotation over time \cite{1992ApJ...400...80H,
  2000ApJ...543..704D, 2002MNRAS.330...35A, 2002ApJ...569L..83A,
  2003MNRAS.341.1179A, 2003MNRAS.346..251O, 2005MNRAS.363..991H,
  2006ApJ...637..214M, 2007MNRAS.375..460W, 2009ApJ...697..293D}. We have
  performed a simulation of an isolated galactic disk hosting a strong bar
  which includes a state-of-the-art model of the interstellar medium and a live dark matter halo. Here we
  show that in this simulation the bar pattern does not slow down over time,
  and instead remains at a stable, constant rate of rotation. This behavior
  has been observed in previous simulations but its explanation has remained
  elusive.\cite{1993AA...268...65F, 2010ApJ...719.1470V}. We propose that the gas phase of the disk and the dark matter halo act in concert to stabilize the bar pattern and prevent it from slowing down or speeding up. This result
  naturally explains why nearly all observed bars rotate quickly\cite{2011MSAIS..18...23C, 2015AA...576A.102A,
2019MNRAS.482.1733G} and is especially relevant for
  our understanding of how the Milky Way arrived at its present day state.

@@

It has long been known that non-axisymmetric features in a stellar disk act to
redistribute mass and angular momentum.\cite{1972MNRAS.157....1L} The
interaction between a bar and a spheroid (either a dark matter halo or stellar
bulge) has received considerable interest.\cite{1984MNRAS.209..729T,
1985MNRAS.213..451W} A stellar disk in isolation seems to always
generate a bar instability,\cite{1971ApJ...168..343H} but the presence of a strong spherical potential (e.g., from a stellar bulge or dark matter halo) acts to stabilize the disk against the bar
instability.\cite{1973ApJ...186..467O, 1976AJ.....81...30H} A number of
numerical simulations have confirmed that a live dark matter halo generically
exerts a negative torque on the bar, causing its pattern speed to decrease and
its length and strength to increase\cite{1992ApJ...400...80H,
2000ApJ...543..704D, 2002MNRAS.330...35A, 2002ApJ...569L..83A,
2003MNRAS.341.1179A, 2003MNRAS.346..251O, 2005MNRAS.363..991H,
2006ApJ...637..214M, 2007MNRAS.375..460W, 2009ApJ...697..293D}. Intuitively,
the bar excites a wake of resonant material in the dark matter halo in a
manner similar to dynamical friction. This wake lags and thus exerts a
negative torque on the bar.

The preeminent expectation that the dark matter halo acts to slow the bar
down conflicts with observational estimates of bar pattern speeds. Bar
rotation rates are typically classified by the dimensionless ratio
\begin{equation}
\Rot = \RCR/R_b\textrm{,}
\end{equation}
where \RCR{} is the radius of corotation and $R_b$ is the length of the
bar.\footnote{The radius of corotation \RCR{} is defined for circular orbits as
the radius at which the orbital frequency is equal to the pattern speed,
$\Omega_p$, of a given non-axisymmetric feature. In a galaxy with a constant
circular velocity $V_c$, it is given by $\RCR = V_c / \Omega_p$.} Galaxies
with $\Rot < 1.4$ are considered ``fast rotators'' while galaxies with $\Rot >
1.4$ are considered ``slow rotators.''\cite{2000ApJ...543..704D} Galaxies with
$\Rot < 1$ are not thought to be stable.\cite{1980AA....81..198C}
Observational estimates of the pattern speeds of bars indicate that nearly all
galaxies have $1 < \Rot < 1.4$.\cite{2011MSAIS..18...23C, 2015AA...576A.102A,
2019MNRAS.482.1733G}

@@

A surface density projection of our simulations is shown in
Fig.~\ref{fig:overview}. The upper three panels show the disk in the N-body
run while the lower three panels show the disk in the SMUGGLE run. Each column
shows the disk $\sim1\,\textrm{Gyr}$ apart in time. There is a large qualitative
difference in the evolution of the bar pattern between the two runs. We see
that in the N-body case, the bar lengthens in time and grows in strength. In
the SMUGGLE case, the bar pattern retains a similar length and strength
between panels.

@@

We can test this interpretation by measuring the angle offset between the halo
wake and the bar. If the wake and the bar are aligned (i.e., there is no angle
offset), then the wake cannot exert a negative torque on the bar. This angle
is plotted in the middle panel of Fig.~\ref{fig:wake}, which shows that the
angle offset is larger in the N-body case than in the SMUGGLE case by about a
factor of two. The left and right panels of Fig.~\ref{fig:wake} show the halo
wake with respect to the location of the bar in the N-body (left) and SMUGGLE
(right) cases.

@@

When the bar slows down, we argue that this induces a larger positive torque
from the gas phase. Only gas within corotation will flow inwards, while gas
outside corotation will flow outwards.\cite{2011MNRAS.415.1027H} Since the
corotation radius is larger for more slowly rotating bars, it follows that
more slowly rotating bars should be more efficient at driving gas inflows and
thus experience a larger positive torque from the gas phase.

We performed an experiment to test this hypothesis by forcing the stellar disk
in the SMUGGLE run to rotate at a constant angular rate and measuring the
torque on the bar by the gas phase at different rotation rates. The result of
this experiment is shown in Fig.~\ref{fig:equil}, which shows that a more
slowly rotating bar experiences a larger positive torque from the gas. We have
therefore shown evidence for an equilibrium mechanism which keeps the pattern
speed of the bar constant, resulting from the complex interplay between the
dark matter halo and the gas phase.

@@

Barred galaxies in cosmological simulations of galaxy formation continue to be
in conflict with observations by producing bars which rotate too slowly.\cite{2017MNRAS.469.1054A, 2019MNRAS.483.2721P,
2021AA...650L..16F} However, the pattern speeds of bars in both cosmological
simulations and the real universe can be affected by environmental processes
not included in our simulation -- e.g., satellite
infall\cite{2011Natur.477..301P}, non-sphericity\cite{2013MNRAS.429.1949A} or
rotation\cite{2013MNRAS.434.1287S, 2014ApJ...783L..18L, 2018MNRAS.476.1331C,
2019MNRAS.488.5788C} in the dark matter halo, or perhaps even the gaseous
circumgalactic medium. Naturally, extending our present work to account for
such effects is a crucial next step in understanding the formation and
evolution of galactic bars.

@@

\begin{figure}[h]%
\centering
\includegraphics[width=0.45\textwidth]{fig/fig2.pdf}
\caption{Various bar properties from the N-body and SMUGGLE runs. The
\textit{upper panel} show the evolution of the pattern speed. As expected, the
bar in the N-body run slows down due to interactions between the bar and the
dark matter halo. However, the bar in the SMUGGLE run does not slow down and
instead remains at a constant pattern speed. The \textit{middle panel} shows
the evolution of the bar length. In the N-body case, the bar lengthens. This
occurs because as the pattern speed drops, bar-like orbits at larger radii are
possible. Stars are captured on these orbits, lengthening the bar. This
process does not occur in the SMUGGLE cases since the bar pattern speed is not
decreasing, and therefore the bar length remains constant. The \textit{lower
panel} shows the torque on the bar by different components. The solid lines
show the torque exerted by the halo in both the N-body and SMUGGLE cases. The
dashed line shows the torque exerted by the gas phase in the SMUGGLE run
(there is no gas in the N-body run). Details on how these properties
are calculated is given in the Methods section.}\label{fig:prop}
\end{figure}

@@

Star formation occurs in cells above a density threshold
($n_{\textrm{th}}=100\,\textrm{cm}^{-3}$) according to
ref.\cite{2003MNRAS.339..289S} with a star-formation efficiency of $\epsilon =
0.01$. Star formation converts gas cells into star particles which represent
single stellar populations. For each star particle, the deposition of energy,
momentum, and mass from stellar winds and supernovae is modeled.
Photo-ionization and radiation pressure are modeled using an approximate
treatment. A more detailed description of this model can be found in the
flagship SMUGGLE paper.\cite{2019MNRAS.489.4233M}

We used the fiducial model parameters, except that we increased the number of
effective neighbors $N_{\textrm{ngb}}$ for the deposition of feedback from
$64$ to $512$. We found that a lower value of $N_{\textrm{ngb}}$ resulted in
inefficient photo-ionization feedback since the photo-ionizing budget had not
been exhausted after deposition into $64$ neighboring cells. We also used an
updated version of SMUGGLE using a new mechanical feedback routine similar to
the one described in ref.\cite{2018MNRAS.480..800H} This updated routine is a
tensor renormalization which ensures linear and angular momentum conservation
to machine precision.

\noindent
{\bf Wind Model}
\\
\noindent

In addition to the SMUGGLE model, we considered a simpler model of the interstellar medium based upon

@@

After generating the gaseous disk in this way, we stitched the gas disk
together with the GALAKOS N-body disk (and bulge and dark matter halo) after
the GALAKOS disk has been allowed to evolve for $1.5\,\textrm{Gyr}$. The
purpose of allowing the GALAKOS disk to evolve first for a short period of
time is to allow for the bar to form unimpacted by the presence of the gas
(which would normally disrupt the formation of the bar). Throughout this work,
we consider $t=0$ for the N-body run to be the time at which we added the gas
phase for the SMUGGLE run (i.e. we ignore the first $1.5\,\textrm{Gyr}$ of
evolution of the N-body disk when the bar is forming). We made one additional
modification when stitching the gas disk together with the N-body disk - we
created a vacuum within the central $4\,\textrm{kpc}$. This vacuum guards
against an initial dramatic infall of gas within the bar region, which we
found to destroy the bar.
